---
title: "Creating_raster_mask"
author: "Denisse Fierro Arcos"
date: "2023-11-18"
output:
  github_document:
    toc: true
    html_preview: false
---

## Introduction

In this notebook, we will show how to create a raster mask that can be used to extract data from Earth System Models. We will the transform this mask into a table. We will use the Exclusive Economic Zone (EEZ) shapefile from the [Flanders Marine Institute (2019), version 11](https://doi.org/10.14284/386). This shapefiles does not contain any overlapping boundaries, so we will create two-dimensional masks that will match the various grids used in the Fish-MIP project.  
  
# Loading libraries
  
```{r libraries}
#Spatial data
library(sf)
library(terra)

#Manipulating and plotting data
library(dplyr)
library(stringr)
library(ggplot2)

#Base map
library(rnaturalearth)
```
  
# Loading EEZ shapefile

```{r}
#Loading shapefile
eez_shp <- read_sf("../../World_EEZ_v11_20191118/eez_short.shp")

#Checking result
ne_countries(returnclass = "sf") |> 
  ggplot()+
  geom_sf()+
  geom_sf(inherit.aes = F, data = eez_shp, aes(fill = SOVEREIGN1), alpha = 0.7)+
  theme_bw()+
  theme(legend.position = "none")
```
  
The EEZs are correctly plotted, now we can move onto creating raster masks. In total, we will create four different masks to match all inputs and outputs used in Fish-MIP.  
  
# Finding location of raster samples
We have a folder containing samples of the raster used in Fish-MIP models and ESMs. We will list all the files contained in that folder.    
  
```{r}
sample_rasters <- list.files("../ESM_Sample_Data/", pattern = ".nc$", full.names = T)
sample_rasters
```
  
We will define a function that will go through each sample file and create a mask.  
  
```{r}
shp_to_raster <- function(shp, raster_path, out_folder){
  #Transforming shapefile to SpatVector class
  shp_terra <- vect(shp)
  #Loading raster
  ras <- rast(raster_path)
  #Rasterise shapefile
  shp_rast <- rasterize(shp_terra, ras, field = "MRGID", background = NA)
  #Create name for mask to be saved from original raster sample
  file_out <- paste0("EEZ_mask", 
                     str_extract(raster_path, "area(_.*nc)", group = 1))
  file_out <- file.path(out_folder, file_out)
  writeCDF(shp_rast, file_out, overwrite = T, varname = "eez", 
           longname = "Economic_Exclusive_Zones", missval = NA)
}
```
  
Applying function to all raster samples.  
  
```{r}
#Ensure output folder exists
out_folder <- "../Outputs/EEZMasks"
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)}

#Applying function creating masks
for(ras in sample_rasters){
  shp_to_raster(eez_shp, ras, out_folder)
}
```
  
# Plotting mask
We will plot one mask to ensure it has been correctly created.  
  
```{r}
ras <- rast("../Outputs/EEZMasks/EEZ_mask_1deg_DBPM.nc")
plot(ras)
```
  
# Creating 2D `csv` mask
We will define a function that we use to create a data frame from the raster masks.  
  
```{r}
mask_df <- function(raster_path){
  df <- rast(raster_path) %>% 
    #Transforming form matrix to data frame
    as.data.frame(xy = T) %>% 
    #Renaming latitude and longitude columns
    rename("lon" = "x", "lat" = "y")
  
  #Create name for mask to be saved from original raster sample
  file_out <- str_replace(raster_path, ".nc", ".csv")
  #Save file
  write_csv(df, file_out) 
}
```
  
Now we will apply the function to all the raster masks.
  
```{r}
raster_masks <- list.files(out_folder, pattern = ".nc$", full.names = T)

#Applying function creating csv masks
for(ras in raster_masks){
  mask_df(ras)}
```
  
We can now use these masks to extract data from any rasters sharing the same grid as the mask.  
  
